---
title: "Ions demo"
author: "Balazs B Ujfalussy"
date: "11/09/2017"
output:
  pdf_document: default
  html_document: default
---

This is a demo for simulating the buildup of the equilibrium potential in a simple neuron. We use the package [deSolve](http://desolve.r-forge.r-project.org) to simulate the differential equations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source('demos/ions_sim.R', chdir=T)
```

Both electric and chemical changes are simulated. The goal is to compare the time course and magnitude of the changes under different conditions.

There are two simulations in this demo:
1: Nernst potential
2: equilibrium potential


## 1. Simulating the Nernst equation

We will simulate the flux of K ions through a simple membrane permeable only to K ions.
Initially there are no electric potential difference between the two sides of the membrane but there is a difference in the ionic concentrations


```{r}
params <- c(gK=gK*2, cm=cm, vi=vi, ve=ve) # parameters of the system. Parameters are: 
# gK: K conductance in mS, 
# cm: membrane capacitance in uF
# vi and ve: the intracellular and extracellular volume in cm3. 
# For more details, see the demos/ions_consts.R file
state <- c(v=0, C.Ki=C.Ki.init, C.Ke=C.Ke.init) # initial state of the system. State variables are:
# v - membrane potential - mV
# C.Ki - intracellular K concentration in mM
# C.Ke - extracellular K concentration in mM

times <-  seq(0,30, by=1/10) # the time axis of the simulations
out <- ode(y = state, times = times, func = sim.Nernst, parms = params) # the function ode solves the system of differential equations
```


Now, we plot the results

```{r, echo=FALSE}

icol <- rainbow(24)[15]
layout(matrix(1:3, 3), 1, c(2,2,3))
par(mar=c(1,5,3,1))
plot(out[,1], out[,2], t="l", col=icol, axes=F, xlab="", ylab="", main="membrane potential (mV)")
axis(2, las=2)
par(mar=c(1,5,3,1))
plot(out[,1], out[,3], t="l", col=icol, axes=F, xlab="", ylab="", main="[K]-intra (mM)")
axis(2, las=2)
par(mar=c(4,5,3,1))
plot(out[,1], out[,4], t="l", col=icol, axes=F, xlab="time (ms)", ylab="", main="[K]-extra (mM)")
axis(2, las=2)
axis(1)
```

## 2. Equilibrium potential

We will simulate the buildup of the resting potantial of a neuron. Initially there are no electric or chemical gradients. At t=0, we switch on the Na/K exchanger, that starts transporting ions. Both Na and K ions are simulated, the permeability is assumed to be constant.

```{r}
# simulation
params <- c(gK=gK, gNa=gNa, cm=cm, vi=vi, ve=ve, I.pump.K=I.pump.K, I.pump.Na=I.pump.Na)# simulation parameters.
# gK, gNa: K and Na conductance in mS, 
# cm: membrane capacitance in uF
# vi and ve: the intracellular and extracellular volume in cm3. 
# I.pump.K, I.pump.Na: the current mediated by the Na/K exchanger in uA
# For more details, see the demos/ions_consts.R file
state <- c(v=0, C.Nai=70, C.Nae=70, C.Ki=65, C.Ke=65)
# initial state of the system. State variables are:
# v - membrane potential - mV
# C.Ki, C.Nai - intracellular K and Na concentration in mM
# C.Ke, C.Nae - extracellular K and Na concentration in mM

times <-  seq(0, 5000000, by=1000) # in ms!
out2 <- ode(y = state, times = times, func = sim.equilibrium, parms = params)
```

Now look at the results

```{r, echo=FALSE}
layout(matrix(1:2, 2), 1, c(4,5))
par(mar=c(1,5,3,1))
plot(out2[,1], out2[,2], t="l", col=1, axes=F, xlab="", ylab="", main="membrane potential (mV)"); axis(2, las=2)
par(mar=c(4,5,3,1))
matplot(out2[,1], out2[,3:6], t="l", col=c(2,2,3,3), lty=c(1,2,1,2), axes=F, xlab="time (s)", ylab="", main="ion concentrations (mM)"); axis(2, las=2)
axis(1, seq(0, 5000000, by= 1000000), seq(0, 5000, by= 1000))
legend('topright', c('[Na] intra', '[Na] extra', '[K] intra', '[K] extra'), col=c(2,2,3,3), lty=c(1,2,1,2), bg=grey(1, alpha=0.75))
```



