---
title: "Analysing place cells"
author: "Balazs B Ujfalussy"
date: "20/11/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a demo for illustrating the activity of place cells. We will analyse a 30 min recording session from a rat running back and forth in a 1.6 m ling linear track, from the laboratory of Gyuri Buzsaki. There are about 120 pyramidal cells recorded simulataneously, half of them have place field in this environment. There are also 20 interneurons. 

There are two exercices related to the dataset: 

* analysing the coding properties of place cells - how reliable they are? How much information they contain?
* decoding the activity of place cells - can you predict the position of the rat from the spike counts of the cells?

The source of the data:
Grosmark, A.D., Long J. and Buzsáki, G (2016). Recordings from hippocampal area CA1, PRE, during and POST novel spatial learning. CRCNS.org <http://dx.doi.org/10.6080/K0862DC5>

Paper related to the dataset:
Grosmark, A.D., and Buzsáki, G. (2016). Diversity in neural firing dynamics supports
both rigid and learned hippocampal sequences. Science 351, 1440–1443.


## Load the dataset and observe place cells

We will read a preprocessed data file `Achilles.RData` - Achilles is the name of the rat. It contains a list named `rat` that stores different variables describing the experiment.
```{r}
load('./Achilles.RData')
```

The `summary()` function tells you what variables are encoded int the list `rat`:
```{r}
summary(rat)
```

* The main variables encoded are the position (`pos`), spike times (`spt`). The position is a matrix of two columns: time in seconds and smoothed 1D position of the animal along the linear track. The spt is also a two colummn matrix, time in seconds and the ID of the cell that emitted the spike. (Cell id's refer to the elctrodes the cell was recorded from, so they do not start from 1...)

* Before and after each run the animal stays at the end for a while to consume reward. The activity of the place cells can depend on the direction of the movement, so up and down runs (left and right, sorry :-)) are treated differently. The variable `iruns.down` and `iruns.up` stores the index of the start and the end of the individual down or up runs indexing the rows of the matrix `pos`.

* The variable `PyrIDs` and `IntIDs` stores the name of the (putative) pyramidal cells and interneurons. 

* `MazeRange` stores the x coordinates associated with the start and the end of the runs. 

The individual variables in the list can be referred by the \$ sign. For example the dimensionality of the position variable can be prompted as `dim(rat$pos)` which returns `r dim(rat$pos)` meaning that this matrix has `r nrow(rat$pos)` rows and `r ncol(rat$pos)` columns.

Now we will load a function that will analyse this data to return the spike counts for each cell on each runs in the function of the (discretised) position.
```{r}
source("PlaceCellFunctions.R")
pos <- rat$pos
spt <- rat$spt
```


We define the spatial discretization in 5 cm.
```{r}
dx <- 0.05 # cm, resolution 
x.breaks <- seq(rat$MazeRange[1], rat$MazeRange[2], by=dx)
x.mids <- round(x.breaks[-1] - dx/2, 3)
```


Next we prepare an array named `act.runs` that contains the spikes of each cell on each trial with spatial resolution dx. Its dimensions are {number of cells} x {distance} x {trials}. The last neuron is not a true neuron, but stores the time (in seconds) the rat spent at each location at each trial. We only analyse the up runs here, but one of the homeworks is to compare the place fields in the two directions. The function `cell.maps.runs` extracts the relevant data for us.

```{r, cache=TRUE}
act.runs <- cell.maps.runs(spt, pos, i.runs=rat$iruns.up, dx=0.05, MazeRange=rat$MazeRange, cell.IDs=rat$PyrIDs)
```


We divide spike count with the occupancy time to get firing rates, and plot the firing rate of all cells in the function of distance.
```{r, fig.cap='Ratemap of all pyramida cells.'}
ratemaps.t <- apply(act.runs[1:120,,], c(1,2), sum)
Tmap <- apply(act.runs[121,,], 1, sum)

ratemaps.all <- t(ratemaps.t) / Tmap
matplot(x.mids, ratemaps.all, t='l', lty=1, col=rainbow(120), xlab='x position (m)', ylab='firing rate (Hz)')
```

Next, plot firing rates of two example neurons estimated on individual runs - we still need to divide with occupancy time!


```{r, fig.cap='Firing rates on individual trials for two neurons. The left cell looks like a classical place cell, the right starts to fire only after the 30th run.'}
par(mfcol=c(1,2)); par(mar=c(4,4,4,4))
image(x.mids, 1:42, act.runs[70,,] / act.runs[121,,], col=topo.colors(24), xlab='x position (m)', ylab='trial', main='', ylim=c(0, 42))
lines(x.mids, ratemaps.all[,70], col=heat.colors(3)[2], lwd=2)
axis(4, c(0, 5, 10), c(0, 5, 10))
mtext('firing rate (Hz)', 4, 2, adj=0)
image(x.mids, 1:42, act.runs[97,,] / act.runs[121,,], col=topo.colors(24), xlab='x position (m)', ylab='trial', main='', ylim=c(0, 42))
lines(x.mids, ratemaps.all[,97], col=heat.colors(3)[2], lwd=2)
axis(4, c(0, 5, 10), c(0, 5, 10))
mtext('firing rate (Hz)', 4, 2, adj=0)
```


Finally plot the ratemap of all active cells - sorted according to the position of the peak.

```{r}
i.cells.active <- which(apply(ratemaps.all, 2, max) > 5)
N.cells.active <- length(i.cells.active)
ratemaps <- ratemaps.all[,i.cells.active]
ii.maxs <- apply(ratemaps, 2, which.max)
sort.peaks <- sort(ii.maxs, ind=T)$ix
```

```{r, fig.cap='Place cells sorted according to their peak firing rate'}
par(mfcol=c(1,2))
matplot(x.mids, ratemaps[,sort.peaks], t='l', lty=1, col=rainbow(60), xlab='x position (m)', ylab='firing rate (Hz)')
image(x.mids, 1:N.cells.active, ratemaps[,sort.peaks], col=topo.colors(24), xlab='x position (m)', ylab='cell')
```

## Homeworks

1. Reliability of the place cells. A place cell is useful if it always works, i.e., it fires action potential whenever the animal is at a certain location. If it only signals at half of the times, it seems to be less informative. Look at the place cells of Figure 2. Which one is a more useful as a place cell? [2p]

2. Look at the activity of other place cells on individual runs. Are they reliable indicator of the animal's location? Can you find examples of reliable and unreliable cells? Can you identify other classes? [4p]

3. Can you come up with a simple metric that measures the reliability of a place cell? Show how it works on two example place cells! [6p]

4. Calculate the reliability metric for the entire population of place cells and investigate whether it correlates with our intuitive sense of reliability! You can get the full points even if the metric does not work, but try to understand why it fails! [4p]

5. Calculate the place fields of the same place cells on the opposite runs. Compare the location of the place fields in the two direction. Can you find cells with the activity  truely depending on the position and independent of the running direction? Can you find cells that are sensitive to the running direction? [4p]

Calculate the information rate (bits/spike) for all place cells, which is defined in Skaggs et al.: An information theoretic approach to deciphering the hippocampal code. 1993 as:

$I = \sum_x r(x)\, P(x) \log \frac{r(x)}{\hat{r}}$

where $r(x)$ is the firing rate of the cell as the function of position, $\hat{r} = \sum_x P(x)\,r(x)$ is the mean firing rate and $P(x)$ is the estimated probability of being at position $x$. $I$ is measured in bits/second (if a base 2 logarithm is taken).

Calculate the above quantity for the place cells. Compare it with their activity map. Do you think that cells with higher information rate are more informative about the location of the animal?

What is the information rate of the two cells shown in Fig. 2? Does the difference in their information rate reflets the striking difference in their behaviour? What other measure would you use to discriminate the left and the right cell in Fig. 2? Implement it and show how it works (or describe why it fails)!


## Homework 2 - decoding


Now we will do static decoding of place cell activity. First, we need the spike patterns and the position in temporal bins of 0.1 second width. Note, that now we analyse the same dataset in temporal windows. Before we analysed it in spatial windows. The function `pop.act.t` will return the population activity during the run sessions in approximately 100 ms time windows. This will take some time (~1 min on my computer), but you will be able to follow the progress of the algorithm.

```{r, cache=TRUE, message=F, echo=T, results='hide'}
all.data <- pop.act.t(spt, pos, i.runs=rat$iruns.up, dt=0.1, cell.IDs=rat$PyrIDs[i.cells.active])
```

```{r}
summary(all.data)
```

The list `all.data` contains two variables: `pos` is a vector, with the $x$ position and `popact` is a matrix with the population activity. Each row in `all.data$popact` is a cell and each column shows the spike counts $\mathbf{s}$ in the given time bin. We have `r length(all.data$pos)` time points.


We prepare two datasets, a training and a test dataset, each containing the spike counts of the `r length(i.cells.active)` active cells and the location of the animal $x$. We will randomly select 2000 time points for training and the rest for testing.

```{r}
L.all <- length(all.data$pos)
L.train <- 2000
L.test <- L.all - L.train

i.train <- sample (1:L.all, L.train)
i.test <- which(!(1:L.all %in% i.train))

data.train <- list(pos=all.data$pos[i.train], popact=all.data$popact[,i.train])
data.test <- list(pos=all.data$pos[i.test], popact=all.data$popact[,i.test])
```

We can learn the relationship between $\mathbf{s}$ and $x$ using the training data. Then we can predict the position $x$ from the spikes $\mathbf{s}$. Use one of the learned techniques to predict the position $x$ from the spikes $\mathbf{s}$ in the test data - `data.test`.
